---
title: |  
    | How to Avoid Overestimating the Burden of Hypertension in Epidemiological Studies? A New Methodology to Account for Within-Person Variability of Blood Pressure
    | <font size="4">  *Manuscript submitted for publication, 2022*  </font> 
subtitle: A working example
author: Edouard Chatignoux, Valérie Olié, Christophe Bonaldi, Clémence Grave, Amélie Gabet, Jacques Blacher
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    <!-- css: style.css  -->
    <!-- code_folding: hide -->
    highlight: tango
    fig_width: 9
    fig_height: 9
    dev: svg
    number_sections: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    fig_caption: TRUE
---



```{r,eval=T,echo=F}

options(dplyr.print_min = 5)
options(dplyr.print_max = 5)
knitr::opts_chunk$set(echo=T, fig.align="center",cache=T,
                      message = FALSE, warnings = FALSE)

## set.seed(10)
## dt_calib<-dt_calib%>%
##   group_by(site,dist,age)%>%
##   mutate(N_I=rpois(lambda=N_I,n=1))%>%
##   rowwise()%>%
##   mutate(N_HC=rpois(lambda=N_HC,n=1))

``` 

# Overview#

This document illustrates the application of a
methodology to take into account within-person blood pressure (BP) variability in
epidemiological studies to avoid bias in the estimation of
the burden of hypertension (HT).


R-package
---------

The material and R code used in this application are part of the 
R-package `BPpack`, available in Github. 
Use the package [devtools](https://github.com/hadley/devtools) to
install the package : 

```{r,echo=T,eval=F}

require(devtools)
install_github("echatignoux/BPpack")

```

After installation, the package can be loaded into R.

```{r,echo=T}

library(BPpack)

```

The `tidyverse`, `magrittr`, `splines` and `survey` packages are also needed 
to run the following code. 



```{r,echo=T}

library(tidyverse)
library(magrittr)
library(splines)
library(survey)

``` 


Data sources
---------

The application of the method is illustrated here on the NHANESIII
data base, available at
(https://wwwn.cdc.gov/nchs/nhanes/nhanes3/default.aspx).

For the purpose of illustration, we restricted the NHANESIII data to
patients having at least two visits and two measures of BP per
visit. We further selected, for each patients, the first two visits
and the last two BP measures per visit, so that the resulting data in
`dt_nhanes` corresponds to an equilibrated design.

The resulting data is stored in a long format, with one line per
patient/visit/measure.

```{r}
dt_nhanes

``` 


Main function
---------

The main function of the `BPpack` package is the `correct_htn`
function, which allows for estimation of the prevalence of
hypertension.

Its arguments are (see `?correct_htn`):

* `form` A formula giving covariates by which HTN prevalence is
   calculated. If `htn` is placed in the left hand side of
   `formula`, then variation of prevalence according to the
   covariates given in the right hand side of are estimated with a
   `glm` (`survey::svyglm` if `surv_des` is not NULL) model with
   quasibinomial distribution. Else, prevalence is tabulated
   according to the covariates given in the right hand side (simple
   tabulation or `survey::svyby` if `surv_des` is not NUL).
*  `subpop` A boolean covariate that defines a subpopulation
   over which to filter the calculation of prevalence. `subpop`
   must be given in a formula form, e.g. ~subpop. 
* `n_samp` The number of posterior sample of the correction
   factor to be used in the estimation. Default (NULL) resumes to
   the maximum number of available posterior samples.
* `data_long` Data frame giving the BP measurements of the
   population under study. Data must be given in a long format,
   e.g. one raw per BP measure (see `dt_nhanes` format).  The data
   frame must have the following columns: "id" Patient identifier;
   "age" Age of the patient in years; "sex" Sex of the patient;
   "tt_htn" Boolean with value `TRUE` if the patient is under anti-hypertensive
   treatment; "visit" : Identifier of the visit; "BP_typ" : type of BP
   prelevement, "dia" for diastolic, "sys" for systolic; "bp" : Value of blood pressure
   measurement
* `surv_des` If applicable, the survey design of the study,
   specifyed with `svydesign` from `survey` package. 
* `correct` Boolean set to `TRUE` (the default) to correct
   prevalence.
* `tresh` A data frame giving the BP thresholds that defines
   hypertension. Default to 140 for systolic BP, 90 for diastolic.
* `glm` Boolean set to `TRUE` (the default) to evaluate the prevalence
of HTN according to the formula specified in  `form` with a `glm`
model with binomial distribution (`svyglm` if a survey design is
specified). Otherwise tabulation of the prevalence of HTN according to the covariates in formula is done. 

# Estimation of the burden of hypertension in NHANESIII data #

## Uncorrected estimates ##

Hypertension is defined within one individual if the mean of its BP
measurements (for diastolic or systolic BP) exceeds a threshold, or if
the patient takes an treatment for HT. 

The crude estimation of the prevalence of hypertension consists in
counting hypertensive patients from the data. This can be done using
the `correct_htn` function:

```{r}
correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            correct = FALSE)
``` 

The sampling design of the NHANESIII survey may be taken into account
using the `surv_des` argument of the function:

```{r}

## First, define the design
d.s <- svydesign(ids=~1,
                 data=dt_nhanes,
                 weights=~samp_weight)

## Then, use it 
correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            surv_des = d.s,
            correct = FALSE)

``` 

It worth saying that in example above, the specification of the survey
design was loose, as applied to data given in long format, whereas it
should have been defined from data in a wide format instead. This
doesn't matter actually, as the `correct_htn` function only keeps the
specification of the survey design, and applies it to data with only
one observation per individual.

This point is confirmed bellow:
```{r}

## First, define the design
dt_nhanes_wide<-
  dt_nhanes%>%
  ungroup()%>%
  mutate(vm=paste("bp",visit,meas,sep="_"))%>%
  select(-visit,-meas)%>%
  spread(vm,bp)
d.s.w <- svydesign(ids=~1,
                 data=dt_nhanes_wide,
                 weights=~samp_weight)

## Then, use it 
correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            surv_des = d.s.w,
            correct = FALSE)

``` 

The `correct_htn` function also allows for estimations of HT
prevalence according to covariates, for example, according to body
mass index (variable `bmi`).

Note that this estimation can be done only in the subpopulation with
non missing value of BMI, so that we should restrict or estimates to
this subpop. This is done using the `subpop` argument of the `correct_htn`.


```{r,fig.height=7,fig.width=10}

## Define boolen for the subpop with not missing bmi
dt_nhanes%<>%mutate(has_bmi=!is.na(bmi))
## Estimate the variation of HT according to bmi with a natural spline of 3 df
bmi_htn<-correct_htn(form=htn ~ ns(bmi,df=3),
            data_long = dt_nhanes,
            subpop=~has_bmi,
            surv_des = d.s,
            correct = FALSE)

## Plot the results
bmi_htn%>%
  mutate_at(vars(htn,low,up),~.x*100)%>%
  ggplot(data=.)+
  aes(bmi,htn,ymin=low,ymax=up)+
  geom_line()+
  geom_ribbon(fill=NA,colour="black",linetype=5)+
  xlab("Body mass index")+ylab("HT prevalence (in %)")

``` 


## Corrected estimates ##

Taking into account the within-person blood pressure (BP) variability in the estimation of
hypertension is simply done passing the `correct` argument from the
`correct_htn` function to `TRUE`. The number of samples from the
posterior distribution used for the estimation is controlled by the
`n_samp` argument. As the estimation time increases with the number of
samples, we'll keep it low (`n_samp=100`) in this illustrative application here.


To illustrate the consequences of neglecting within-person blood
pressure variability in the estimates, we estimated HT prevalence 
using the two visits from NHANESIII, or only the first or the second. 

```{r,echo=F}



dt_nhanes_v1<-dt_nhanes%>%filter(visit==1)
dt_nhanes_v2<-dt_nhanes%>%filter(visit==2)

htn_cor<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)

htn_cor_v1<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v1,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)

htn_cor_v2<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v2,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)

htn_uncor<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            surv_des = d.s,
            n_samp = 1,
            correct = FALSE)

htn_uncor_v1<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v1,
            surv_des = d.s,
            n_samp = 1,
            correct = FALSE)

htn_uncor_v2<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v2,
            surv_des = d.s,
            n_samp = 1,
            correct = FALSE)


``` 



```{r,echo=T,eval=F}

dt_nhanes_v1<-dt_nhanes%>%filter(visit==1)
dt_nhanes_v2<-dt_nhanes%>%filter(visit==2)

htn_cor<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)

htn_cor_v1<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v1,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)

htn_cor_v2<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v2,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)

htn_uncor<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes,
            surv_des = d.s,
            n_samp = 1,
            correct = FALSE)

htn_uncor_v1<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v1,
            surv_des = d.s,
            n_samp = 1,
            correct = FALSE)

htn_uncor_v2<-correct_htn(form=htn ~ 1,
            data_long = dt_nhanes_v2,
            surv_des = d.s,
            n_samp = 1,
            correct = FALSE)


``` 


The results are reported in the table below:

```{r, echo=F}
library(kableExtra)

compare_htn<-
  htn_cor%>%mutate(visit="all",correct=TRUE)%>%
  bind_rows(htn_uncor%>%mutate(visit="all",correct=FALSE))%>%
  bind_rows(htn_cor_v1%>%mutate(visit="v1",correct=TRUE))%>%
  bind_rows(htn_uncor_v1%>%mutate(visit="v1",correct=FALSE))%>%
  bind_rows(htn_cor_v2%>%mutate(visit="v2",correct=TRUE))%>%
  bind_rows(htn_uncor_v2%>%mutate(visit="v2",correct=FALSE))%>%
  select(visit,correct,everything())

compare_htn%>%
  mutate_at(vars(htn,low,up,se),~format(round(.x*100,2),nsmall=2))%>%
  mutate(htn=paste0(htn," [",low," - ",up,"]"))%>%
  select(-low,-up,-se)%>%
  mutate(correct=factor(correct,labels=c("Un-corrected HT","Corrected HT")))%>%
  mutate(visit=fct_recode(factor(visit),All="all","Visit 1"="v1","Visit 2"="v2"))%>%
  spread(correct,htn)%>%
  rename(Visit=visit)%>%
  kable(caption="Corrected and un-corrected estimates of HT according to the visits used",escape=FALSE,booktabs = TRUE) %>%
  kable_styling("hover", full_width = F)

  
``` 

The table shows the importance of accounting for BP variability when
estimating HT prevalence. If no correction is performed, HTN estimated
with visit 1 or visit 2 data only are larger than HTN estimated with
the two visits. This latter however also exhibits some bias, corrected
HTN (i.e. HT prevalence estimated with BP collected over an infinite
number visits and measures in the population) being 9% lower than
uncorrected HTN. As expected, the overestimation of HTN is higher
(~15%) when only one visit is used for the estimation.

Some differences however still remains between visit 1 and visit 2
estimates, probably revealing differences in conditions under which BP was
measured in the two visits, including a possible white coat
effect. Without additional information on the reasons for those
differences however, there is no indication about which visit to
favor and no obvious additional correction to apply. 


```{r,eval=F, echo=F}

compare_htn%>%
  select(htn,visit,correct)%>%
  spread(correct,htn)%>%
  mutate(dif = 100*(`TRUE`-`FALSE`)/`FALSE`)

``` 


Similarly, `correct_htn` can be used to estimating the level of
corrected HT prevalence at different BMI levels with the same syntax
as in the previous section, but specifying `correct = TRUE` in the
arguments.

The corrected curve (in blue) is superimposed with the previous uncorrected
curve (in red).


```{r}
bmi_htn_correct<-correct_htn(form=htn ~ ns(bmi,df=3),
            data_long = dt_nhanes,
            subpop=~has_bmi,
            surv_des = d.s,
            n_samp = 100,
            correct = TRUE)
```


```{r,fig.height=7,fig.width=10, warnings = F, messages = F, echo=F}
## Plot the results
options(warn=-1)
bmi_htn%>%
    mutate(Corrected=FALSE,
           low=NA,up=NA)%>%
    bind_rows(bmi_htn_correct%>%
              mutate(Corrected=TRUE))%>%
    mutate_at(vars(htn,low,up),~.x*100)%>%
    ggplot(data=.)+
    aes(bmi,htn,ymin=low,ymax=up,colour=Corrected)+
    geom_line()+
    geom_line(aes(bmi,up),linetype=5)+
    geom_line(aes(bmi,low),linetype=5)+
    xlab("Body mass index")+ylab("HT prevalence (in %)")+
    theme(legend.position="bottom")


``` 


<!-- Local Variables: -->
<!-- ispell-check-comments: exclusive -->
<!-- ispell-local-dictionary: "english" -->
<!-- End: -->
